Equalizer Programming Guide

### 2009-04-20 Developer/ProgrammingGuide/paper.tex
Programming Guide: update layout API
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.6 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    \hline
  \end{tabularx}}
\clearpage

@@

Canvases provide a convenient way to configure projection surfaces. A
canvas uses a layout, which describes the logical views. The usage of
canvases and layouts is optional. Typically, each desktop window uses
one canvas.

@@

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
  {\caption{A display wall using one canvas and layout with six segments
    and two views.}}
\end{figure}

@@

\textsf{eqPly} uses static distributed objects to distribute the model
loaded by the application. This approach can be easily extended to
versioned objects to support dynamic data modifications.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
The identifier of the model's root object of this distributed hierarchy
is passed as part of the \textsf{InitData} for the default model, or as
part of the \textsf{View} for each logical view. It is used on the
render clients to map the model when it is needed for
rendering. \fig{fModelDist} shows the UML hierarchy of the model and
distribution classes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[ht!]\center
  \vspace{-2ex}
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

%\begin{figure}[ht!]\center
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed. \fig{fConfigInit} shows a
sequence diagram of the config initialization.

@@

For layout and model selection, \textsf{eqPly} maintains an active
view. The identifier of the active view is stored in the frame data,
which is used by the render client to highlight it using a different
background color. The active view can be selected by clicking into a
view, or by cycling through all the views using a keyboard shortcut:

{\footnotesize\begin{lstlisting}
bool Config::handleEvent( const eq::ConfigEvent* event )
{
    switch( event->data.type )
    {
        ...
        case eq::Event::POINTER_BUTTON_PRESS:
            _frameData.setCurrentViewID( event->data.context.view.id );
            return true;
        ...
    }
}

bool Config::_handleKeyEvent( const eq::KeyEvent& event )
{
    switch( event.key )
    {
        ...
        case 'v':
        case 'V':
        {
            NextViewFinder finder( _frameData.getCurrentViewID( ));
            accept( finder );
            const eq::View* view = finder.getResult();
            if( view )
                _frameData.setCurrentViewID( view->getID( ));
            else
                _frameData.setCurrentViewID( EQ_ID_INVALID );
            return true;
        }
        ...
    }
}
\end{lstlisting}}

The model of the active view can be changed using a keyboard
shortcut. The model is view-specific, and therefore the model identifier
for each view is stored on the view, which is used to retrieve the model
on the render clients. View-specific data is not limited to a
model. Applications can choose to make any application-specific data
view-specific, e.g., cameras, rendering modes or annotations. A view is
a generic concept for an application-specific view on data,
\textsf{eqPly} is using different models to illustrate the concept:

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-8ex}\includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination view of a DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

To draw the model, a helper class for view frustum culling is set up
using the view frustum from Equalizer and the camera position from the
frame data. The frustum helper computes the six frustum planes from the
projection and modelView matrices. During rendering, the bounding
spheres of the model are tested against these planes to determine the
visibility with the frustum.

@@

@@

Due to the entity (de-)-initialization and the potential need to
initialize view-specific data, e.g., a model, a layout switch is
relatively expensive and will stall the application for about a second.

>>>>

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.7 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
    \hline
  \end{tabularx}}
\clearpage

@@

Canvases provide a convenient way to configure projection surfaces. A
canvas uses layouts, which describe logical views. The usage of canvases
and layouts is optional. Typically, each desktop window uses one canvas.

@@

A canvas has one or more layouts. One of the layouts is the active
layout, that is, this set of views is currently used for rendering. It
is possible to specify \textsf{OFF} as a layout, which deactivates the
canvas. It is possible to use the same layout on different canvases.

@@

A view might have an observer, in which case its frustum is tracked by
this observer.

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
  {\caption{A display wall using a six-segment canvas with a two-view layout.}}
\end{figure}

\subsection{Observers}

An observer represents an actor looking at multiple views. It has a head
matrix, defining its position and orientation within the world, and an
eye separation. Typically, a configuration has one
observer. Configurations with multiple observers are used if multiple,
head-tracked users are in the same configuration session, e.g., a
non-tracked control host with two tracked head-mounted displays.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
\textsf{eqPly} uses static distributed objects to distribute the model
loaded by the application. This approach can be easily extended to
versioned objects to support dynamic data modifications.

@@

The identifier of the model's root object of this distributed hierarchy
is passed as part of the \textsf{InitData} for the default model, or as
part of the \textsf{View} for each logical view. It is used on the
render clients to map the model when it is needed for
rendering. \fig{fModelDist} shows the UML hierarchy of the model and
distribution classes.

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\label{fObjects}Scene Graph Distribution}}
\end{figure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

%\begin{figure}[ht!]\center
\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\label{fConfigInit}Config Initialization Sequence}}
  \vspace{-4ex}
\end{wrapfigure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed. \fig{fConfigInit} shows a
sequence diagram of the config initialization.

@@

For layout and model selection, \textsf{eqPly} maintains an active view
and canvas. The identifier of the active view is stored in the frame
data, which is used by the render client to highlight it using a
different background color. The active view can be selected by clicking
into a view, or by cycling through all the views using a keyboard
shortcut.

The model of the active view can be changed using a keyboard
shortcut. The model is view-specific, and therefore the model identifier
for each view is stored on the view, which is used to retrieve the model
on the render clients.

View-specific data is not limited to a model. Applications can choose to
make any application-specific data view-specific, e.g., cameras,
rendering modes or annotations. A view is a generic concept for an
application-specific view on data, \textsf{eqPly} is using different
models to illustrate the concept:

@@

After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination view of a DB compound}}
\end{wrapfigure}
To draw the model, a helper class for view frustum culling is set up
using the view frustum from Equalizer and the camera position from the
frame data. The frustum helper computes the six frustum planes from the
projection and modelView matrices. During rendering, the bounding
spheres of the model are tested against these planes to determine the
visibility with the frustum.

@@

An Equalizer configuration has a vector of observers. Each observer has
its own head matrix and eye separation for tracking. In \textsf{eqPly},
all observers are updated with the same head matrix. A typical
application would assign different tracking data to different observers.

@@

Due to the entity (de-)-initialization and the potential need to
initialize view-specific data, e.g., a model, a layout switch is
relatively expensive and will stall \textsf{eqPly} for about a second.
